\Large \bf Apropos Replacement : Development of a full text search tool for man pages

### 2012-02-29 apropos_replacement.tex
Remove endnote
More fixes

\documentclass[letterpaper,twocolumn,10pt]{article}
\usepackage{usenix,epsfig,endnotes}
\usepackage{listings}
\usepackage{color}
\usepackage{times}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
%\usepackage[pdftex]{graphicx}
\lstset{ %
language=sh,
basicstyle=\footnotesize,
numbers=none,
backgroundcolor=\color{white},
showspaces=false,
showstringspaces=false,
showtabs=false,
frame=single,
tabsize=2,
captionpos=b,
breaklines=true,
breakatwhitespace=false,
title=\lstname
}

@@

%for single author (just remove % characters)
\author{
{\rm Abhinav Upadhyay}\\
$<$er.abhinav.upadhyay@gmail.com$>$
\and
{\rm Joerg Sonnenberger}\\
$<$joerg@NetBSD.org$>$
} % end author

@@

This paper discusses a new implementation of \textit{apropos} done by
Abhinav Upadhyay and mentored by Joerg Sonnenberger as part of
Google Summer of Code 2011, for The NetBSD Foundation. The goal of this project
was to replace the
conventional apropos in NetBSD with a modern version which supports full text
searches. \\

@@

{\tt \small
\lstset{language = XML, caption={No results for free form queries}}
\begin{lstlisting}
$ apropos ``add new user''
add new user: nothing appropriate

$ apropos ``get process status''
get process status: nothing appropriate

$ apropos ``termcap database''
termcap database: nothing appropriate
\end{lstlisting}
}

@@

{\tt \small
\lstset{language = XML, caption={No support for stems or words with same root},}
\begin{lstlisting}
$ apropos ``make directories''
mkdir (1) - make directories

$ apropos ``make directory''
make directory: nothing appropriate

$ apropos ``upgrading package''
pkg_add (1) - a utility for installing and
upgrading software package distributions

$ apropos ``upgrading packages''
upgrading packages: nothing appropriate
\end{lstlisting}
}

@@

{\tt \small
\lstset{language = sh, caption={No spelling correction},}
\begin{lstlisting}
$ apropos ``copy strings''
stpcpy, stpncpy, strcpy, strncpy (3) - copy
strings

$ apropos ``coppy strings''
coppy strings: nothing appropriate
\end{lstlisting}
}

@@

\subsection{Other Problems}
Apart from the search related problems there are a few issues related to
the way man pages are handled in NetBSD. The different aliases of the man pages
are stored on the file system in the from of hard (or symbolic) links and these
have to be specified in the makefiles explicitly using the MLINKS mechanism. This
approach works fine but it is a mess from maintenance point of view.
It should be possible to fix this by utilizing the index already built and
maintained by \textit{apropos}, but yet again, the simplistic implementation of
\textit{apropos} does not leave any room for improvement.

\section{Solution: A new apropos}
This project proposes a very simple and straightforward solution to solve these
problems and make man pages more usable. The idea is to exploit the inherent
semantics and structure of the man pages. The \textit{mdoc(7)} [21] and
\textit{man(7)} [16] macros used for writing man pages are semantically quite
rich, those macros combined with the context of the surrounding section of the
man page in which they have been used, can act
as an indicator of the relevancy of the content. For example, the
{\tt \small
.Nm
}
macro when used inside of the
{\tt \small
NAME
}
section holds more significance than if used inside of any other section. Similarly the description of various \textit{errno} values in the
\textit{intro(2)} man page has more important information than the mention of
those \textit{errno} values in the
{\tt \small
RETURN VALUES
}
section of various man pages from section 3. \\

In essence, it should be possible
to parse and extract the structurally meaningful data
out of man pages and index them in the form of an inverted index [4]. The index
combined with a ranking algorithm which exploits the structured nature of the
data to give back more relevant results to the user, can be used to build a
high quality search interface for manual pages.
Such a search tool will solve many of the problems associated with the conventional \textit{apropos(1)} as noted
before and discussed as follows: \\

\begin{description}
\item[Free Form Queries] \hfill \\
The users can express their queries in more natural language form as the searches
are no longer limited to the the NAME section.
For example, queries like \textit{``installing new software package''} may
now produce relevant results.
\end{description}
\begin{description}
\item[Basic Natural Language Processing Support] \hfill \\
When parsing the man pages and building the index, it is also possible to
pre-process the tokens extracted from the man pages to support some of the very
basic natural language processing functionalities. In this implementation, the
Porter stemming algorithm [5] has been used to
reduce the individual tokens extracted from the man pages to their root words.
This enables support for more flexible searches. For example both
\textit{``Installing new packages''} and \textit{``install new package''} will
return same results. \\

Similarly along with the inverted index, a dictionary of the keywords frequently
occurring in the corpus of man pages is also built and used to support
spelling suggestion.
\end{description}
\begin{description}
\item[Bookkeeping of man page metadata] \hfill \\
In this implementation, additional metadata related to the man pages, for example
an md5 hash, the device id, inode number and modification time
of the man page files are indexed and stored.
This allows for very fast and hassle free update of the index as new man pages are installed or the old ones are updated. \\

Similarly a separate index of all the man page aliases is stored and maintained.
This provides an option to get rid of all the hard or symbolic links of the man
pages scattered throughout the filesystem, and also for clean up of the MLINKS
mess in the makefiles.
\end{description}

\subsection{Tools Used}
The two main operations that are critical for this project are parsing of the
man pages and building of an index of the
data obtained from the parsing process. These are highly specific problems, and
solving them from scratch would be a separate project in its own right,
therefore this project tried to avoid reinventing the wheel and used existing
tools and libraries which are battle tested and very successful.
\textit{libmandoc} [6] from \textit{mdocml} [18] project has been used for
parsing the man pages and \textit{sqlite} [7] for indexing and storing the parsed
data on the file system.

@@

In the second pass \textit{makemandb} generates md5 checksum of the files obtained from the previous stage and for each of these checksums, it checks in
the db, whether this md5 checksum already exists in the db or not. If the md5 checksum is already present in the database, then this was just a false alarm and \textit{makemandb} does not need to parse the file, it simply updates the metadata of that file in the database. In case the md5 checksum of a file does not already exist in the database, then that means the file is either a newly installed man page or an older copy of a man page was updated,
\textit{makemandb} feeds such man pages to the parser and extracts interesting
data out of them, later on storing the data in the FTS table in the database.
\end{description}

\begin{description}
\item[Database Schema] \hfill \\
During the initial stage of the project, the structure of the database was kept
simple. The FTS table consisted of 3 columns:
{\tt
name, name\_desc, desc
}
The
{\tt
name
}
and
{\tt
name\_desc
}
columns stored the name of the man page and the one line description from the
{\tt
NAME
}
section respectively, while the
{\tt
desc
}
column stored the content from the rest of the sections. This approach worked
quite well for a small set of man pages (for example the 4000 or so man pages in
the base set of NetBSD). However, as the number of documents were increased (for example including man pages from pkgsrc), the quality of search started deteriorating. This was mainly because bulk of the parsed data was being stored
in the same column in the database and the ranking algorithm had no way of
identifying more relevant results except from identifying whether the match was found in the {\tt NAME} section or any other section. \\

To rectify this, first there was an attempt to try out better ranking schemes
(the various ranking schemes that were tried out during the development of the
project are discussed later). Better ranking schemes required storing precomputed
term weights [19] on the file system, this almost doubled the storage requirements, therefore such ranking schemes were discarded later on, and the
database schema was decomposed further into more columns. The decomposition into
more columns was done to represent the most commonly occurring sections in usual Unix man pages, for example
{\tt NAME, DESCRIPTION, LIBRARY, RETURN VALUES, EXIT STATUS, ERRORS}
etc. Such a decomposition allowed for an elaborate ranking scheme, which could
assign different weights to each of these columns, based on their relevancy. The
final schema of the database is shown in {\bf Listing 5}
\end{description}

@@

Another key aspect of any search application is to filter out or avoid stop-words
. Words like \textit{``the'', ``a'', ``an'', ``'this''} etc. come under the
category of stop-words, these are the words which are \emph{very} commonly used
in the
language. Stop-words usually skew the search results because of their sheer
frequency in the corpus. The usual practice in the information retrieval world
is to filter such stop-words while building the index, however, in case of this
project, it proved to be non-trivial within the given time constraints.
Therefore it was decided that, rather than eliminating stop-words from the index,
a reasonable trade off can be made by filtering out any stop-words from the
user's search query and querying the database for only the keywords which do
not come under the category of stop-words. \\

@@

  COLUMN NAME       DESCRIPTION
  1. section        Section number
  2. name           The name of the page  
  3. name_desc      Short description from
                    NAME section  
  4. desc           Contains the content
                    from DESCRIPTION
                    section and any other
                    section which is not
                    stored in a separate
                    column.
  5. lib            The LIBRARY section
  6. return_vals    RETURN VALUES section
  7. env            ENVIRONMENT section
  8. files          FILES section
  9. exit_status    EXIT STATUS section
  10. diagnostics   DIAGNOSTICS section
  11. errors        ERRORS section
  12. md5_hash      An md5 hash of the
                    contents of the man
                    page (UNIQUE).
(2) mandb_meta:
    This table maintains essential
    metadata of all the indexed
    man page files.

@@

The implementation of the above scheme proved to be \emph{very} slow. It required
too many computations and these were required to be done for each
search result separately to compute its relevance weight. In a state of the
art search application, the index would usually store precomputed \textit{tf-idf}
weights on the file system, in which case the overhead of computing weights of
each individual search result would require very less work, leading to better
performance. An attempt was made to compute the \textit{tf-idf} weights while
indexing the man pages and store them in a separate table. This lead to high
quality search results but it also brought up a serious issue. Storing
precomputed weights on the file system, more than doubled the storage
requirements of the database index. As a concrete example, it took close
to $90 MB$ to index about $8000$ documents. Without the precomputed weights, 
the database size requirement was close to $45 MB$ for the same number of pages.
\end{description}

@@

\section{Results}
This section shows results of some of the sample queries on new \textit{apropos}
to demonstrate how it performs as compared to the traditional version.
\endnote{Although the new \textit{apropos} by default returns 10
results for a query but in the listings the output has been
snipped to save space
} \\

{\tt \small
\lstset{language = XML, caption={Add new user},}
\begin{lstlisting}
$ apropos ``add new user''
ssh-add(1)      adds private key identities to
the authentication agent...on the command line.
If any file requires a passphrase, ssh-add asks
for the passphrase from the user. The
passphrase is read from the user's tty. ssh-add
retries the last passphrase if multiple
identity files are given...

@@

{\tt \small
\lstset{language = XML, caption={make directory},}
\begin{lstlisting}
$ apropos ``make directory''
make(1) maintain program dependencies
...CURDIR A path to the directory where make
was executed. Refer to the description of PWD
for more details. MAKE The name that make was
executed with argv[0].
For compatibility make also sets .MAKE with the
same value. The...

@@

{\tt \small
\lstset{language = XML, caption={signal number to string},}
\begin{lstlisting}
$ apropos ``signal number to string''
psignal(3)      system signal messages
...the signal number is not recognized
sigaction(2) , the string Unknown signal
is produced. The psiginfo function produces
the same output as the psignal function,
only it uses the signal number information
from the si argument. The message strings can...

@@

\theendnotes

>>>>

\documentclass[letterpaper,twocolumn,10pt]{article}
\usepackage{usenix,epsfig,endnotes}
\usepackage{listings}
\usepackage{color}
\usepackage{times}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage{textcomp}
%\usepackage[pdftex]{graphicx}
\lstset{ %
language=sh,
basicstyle=\footnotesize\ttfamily,
numbers=none,
backgroundcolor=\color{white},
showspaces=false,
showstringspaces=false,
showtabs=false,
frame=single,
tabsize=2,
captionpos=b,
breaklines=true,
breakatwhitespace=false,
upquote=true,
escapechar=\%,
title=\lstname
}

@@

%for single author (just remove % characters)
\author{
{\rm Abhinav Upadhyay}\\
$<$er.abhinav.upadhyay@gmail.com$>$
\and
{\rm J\"org Sonnenberger}\\
$<$joerg@NetBSD.org$>$
} % end author

@@

This paper discusses a new implementation of \textit{apropos} done by
Abhinav Upadhyay and mentored by J\"org Sonnenberger as part of
Google Summer of Code 2011, for The NetBSD Foundation. The goal of this project
was to replace the
conventional apropos in NetBSD with a modern version which supports full text
searches. \\

@@

{\small
\lstset{language = XML, caption={No results for free form queries}}
\begin{lstlisting}
$ apropos %\normalfont\upshape``%add new user%\normalfont\upshape''%
add new user: nothing appropriate

$ apropos %\normalfont\upshape``%get process status%\normalfont\upshape''%
get process status: nothing appropriate

$ apropos %\normalfont\upshape``%termcap database%\normalfont\upshape''%
termcap database: nothing appropriate
\end{lstlisting}
}

@@

{\small
\lstset{language = XML, caption={No support for stems or words with same root},}
\begin{lstlisting}
$ apropos %\normalfont\upshape``%make directories%\normalfont\upshape"%
mkdir (1) - make directories

$ apropos %\normalfont\upshape"%make directory%\normalfont\upshape"%
make directory: nothing appropriate

$ apropos %\normalfont\upshape``%upgrading package%\normalfont\upshape''%
pkg_add (1) - a utility for installing and
upgrading software package distributions

$ apropos %\normalfont\upshape``%upgrading packages%\normalfont\upshape''%
upgrading packages: nothing appropriate
\end{lstlisting}
}

@@

{\tt \small
\lstset{language = sh, caption={No spelling correction},}
\begin{lstlisting}
$ apropos %\normalfont\upshape``%copy strings%\normalfont\upshape''%
stpcpy, stpncpy, strcpy, strncpy (3) - copy
strings

$ apropos %\normalfont\upshape``%coppy strings%\normalfont\upshape''%
coppy strings: nothing appropriate
\end{lstlisting}
}

@@

\subsection{Other Problems}
Apart from the search related problems there are a few issues related to
the way man pages are handled in NetBSD. The different aliases of the man pages
are stored on the file system in the from of hard (or symbolic) links and these
have to be specified in the makefiles explicitly using the {\tt MLINKS}
mechanism. This
approach works fine but it is a mess from maintenance point of view.
It should be possible to fix this by utilizing the index already built and
maintained by \textit{apropos}, but yet again, the simplistic implementation of
\textit{apropos} does not leave any room for improvement.

\section{Solution: A new apropos}
This project proposes a very simple and straightforward solution to solve these
problems and make man pages more usable. The idea is to exploit the inherent
semantics and structure of the man pages. The \textit{mdoc(7)} [21] and
\textit{man(7)} [16] macros used for writing man pages are semantically quite
rich, those macros combined with the context of the surrounding section of the
man page in which they have been used, can act
as an indicator of the relevancy of the content. For example, the
{\tt \small
.Nm
}
macro when used inside of the
{\tt \small
NAME
}
section holds more significance than if it is used elsewhere. Similarly the description of various \textit{errno} values in the
\textit{intro(2)} man page has more important information than the mention of
those \textit{errno} values in the
{\tt \small
RETURN VALUES
}
section of various man pages from section 3. \\

In essence, it should be possible
to parse and extract the structurally meaningful data
out of man pages and index them in the form of an inverted index [4]. The index
can be combined with a ranking algorithm which exploits the structured nature of
the
data to give more relevant results to the user, and to build a
high quality search interface for manual pages.
Such a search tool will solve many of the problems associated with the conventional \textit{apropos(1)} as noted
before and discussed as follows: \\

\begin{description}
\item[Free Form Queries] \hfill \\
The users can express their queries in more natural language form as the searches
are no longer limited to the the NAME section.
For example, queries like \textit{``installing new software package''} may
now produce relevant results.
\end{description}
\begin{description}
\item[Basic Natural Language Processing Support] \hfill \\
When parsing the man pages and building the index, tokens are preprocessed after
extraction from the man pages to support some of the very
basic natural language processing functionalities. In this implementation, the
Porter stemming algorithm [5] has been used to
reduce the individual tokens extracted from the man pages to their root words.
This enables support for more flexible searches. For example both
\textit{``Installing new packages''} and \textit{``install new package''} will
return same results. \\

Similarly along with the inverted index, a dictionary of the keywords frequently
occurring in the corpus of man pages is also built and used to support
spelling suggestion.
\end{description}
\begin{description}
\item[Bookkeeping of man page metadata] \hfill \\
In this implementation, additional metadata related to the man pages, for example
an MD5 hash, the device id, inode number and modification time
of the man page files are kept.
This allows fast incremental update of the index as new man pages are installed or the old ones are updated. \\

Similarly a separate index of all the man page aliases is stored and maintained.
This provides an option to get rid of all the hard or symbolic links of the man
pages scattered throughout the filesystem, and also for clean up of the
{\tt MLINKS} mess in the makefiles.
\end{description}

\subsection{Tools Used}
The two main operations that are critical for this project are parsing of the
man pages and building of an index of the
data obtained from the parsing process. These are highly specific problems, and
solving them from scratch would be a separate project in its own right,
therefore this project tried to avoid reinventing the wheel and used existing
tools and libraries which are battle tested and very successful.
\textit{libmandoc} [6] from the \textit{mdocml} [18] project has been used for
parsing the man pages and \textit{sqlite} [7] for indexing and storing the parsed
data on the file system.

@@

In the second pass \textit{makemandb} generates MD5 checksum of the files obtained from the previous stage and for each of these checksums, it checks in
the db, whether this MD5 checksum already exists in the db or not. If the MD5 checksum is already present in the database, then this was just a false alarm and \textit{makemandb} does not need to parse the file, it simply updates the metadata of that file in the database. In case the MD5 checksum of a file does not already exist in the database, then that means the file is either a newly installed man page or an older copy of a man page was updated,
\textit{makemandb} feeds such man pages to the parser and extracts interesting
data out of them, later on storing the data in the FTS table in the database.
\end{description}

\begin{description}
\item[Database Schema] \hfill \\
During the initial stage of the project, the structure of the database was kept
simple. The FTS table consisted of 3 columns:
{\tt name}, {\tt name\_desc}, and {\tt desc}.
The
{\tt
name
}
and
{\tt
name\_desc
}
columns stored the name of the man page and the one line description from the
{\tt
NAME
}
section respectively, while the
{\tt
desc
}
column stored the content from the rest of the sections. This approach worked
quite well for a small set of man pages (for example the 4000 or so man pages in
the base set of NetBSD). However, as the number of documents were increased (for example including man pages from pkgsrc), the quality of search started deteriorating. This was mainly because bulk of the parsed data was being stored
in the same column in the database and the ranking algorithm had no way of
identifying more relevant results except from identifying whether the match was found in the {\tt NAME} section or any other section. \\

To rectify this, first there was an attempt to try out better ranking schemes
(the various ranking schemes that were tried out during the development of the
project are discussed later). Better ranking schemes required storing precomputed
term weights [19] on the file system, this almost doubled the storage requirements, therefore such ranking schemes were discarded later on, and the
database schema was decomposed further into more columns. The decomposition into
more columns was done to represent the most commonly occurring sections in usual Unix man pages, for example
{\tt NAME}, {\tt DESCRIPTION}, {\tt LIBRARY}, {\tt RETURN VALUES},
{\tt EXIT STATUS}, {\tt ERRORS}
etc. Such a decomposition allowed for an elaborate ranking scheme, which could
assign different weights to each of these columns, based on their relevancy. The
final schema of the database is shown in {\bf Listing 5}
\end{description}

@@

Another key aspect of any search application is to filter out or avoid stop-words
. Words like \textit{``the'', ``a'', ``an'', ``this''} etc. come under the
category of stop-words, these are the words which are \emph{very} commonly used
in the
language. Stop-words usually skew the search results because of their sheer
frequency in the corpus. The usual practice in the information retrieval world
is to filter such stop-words while building the index, however, in case of this
project, it proved to be non-trivial within the given time constraints.
Therefore it was decided that, rather than eliminating stop-words from the index,
a reasonable trade off can be made by filtering out any stop-words from the
user's search query and querying the database for only the keywords which do
not come under the category of stop-words. \\

@@

  COLUMN NAME       DESCRIPTION
  1. section        Section number
  2. name           The name of the page  
  3. name_desc      Short description from
                    NAME section  
  4. desc           Contains the content
                    from DESCRIPTION
                    section and any other
                    section which is not
                    stored in a separate
                    column.
  5. lib            The LIBRARY section
  6. return_vals    RETURN VALUES section
  7. env            ENVIRONMENT section
  8. files          FILES section
  9. exit_status    EXIT STATUS section
  10. diagnostics   DIAGNOSTICS section
  11. errors        ERRORS section
  12. md5_hash      An MD5 hash of the
                    contents of the man
                    page (UNIQUE).
(2) mandb_meta:
    This table maintains essential
    metadata of all the indexed
    man page files.

@@

The implementation of the above scheme proved to be \emph{very} slow. It required
too many computations and these were required to be done for each
search result separately to compute its relevance weight. In a state of the
art search application, the index would usually store precomputed \textit{tf-idf}
weights on the file system, in which case the overhead of computing weights of
each individual search result would require very less work, leading to better
performance. An attempt was made to compute the \textit{tf-idf} weights while
indexing the man pages and store them in a separate table. This lead to high
quality search results but it also brought up a serious issue. Storing
precomputed weights on the file system, more than doubled the storage
requirements of the database index. As a concrete example, it took close
to $90$ MB to index about $8000$ documents. Without the precomputed weights, 
the database size requirement was close to $45$ MB for the same number of pages.
\end{description}

@@

\section{Results}
This section shows results of some of the sample queries on new \textit{apropos}
to demonstrate how it performs as compared to the traditional version. \\
%\footnote{Although the new \textit{apropos} by default returns 10
%results for a query but in the listings the output has been
%snipped to save space
%} \\

{ \small
\lstset{language = XML, caption={Add new user},}
\begin{lstlisting}
$ apropos %\normalfont\upshape``%add new user%\normalfont\upshape''%
ssh-add(1)      adds private key identities to
the authentication agent...on the command line.
If any file requires a passphrase, ssh-add asks
for the passphrase from the user. The
passphrase is read from the user's tty. ssh-add
retries the last passphrase if multiple
identity files are given...

@@

{\tt \small
\lstset{language = XML, caption={make directory},}
\begin{lstlisting}
$ apropos %\normalfont\upshape``%make directory%\normalfont\upshape''%
make(1) maintain program dependencies
...CURDIR A path to the directory where make
was executed. Refer to the description of PWD
for more details. MAKE The name that make was
executed with argv[0].
For compatibility make also sets .MAKE with the
same value. The...

@@

{\tt \small
\lstset{language = XML, caption={signal number to string},}
\begin{lstlisting}
$ apropos %\normalfont\upshape``%signal number to string%\normalfont\upshape''%
psignal(3)      system signal messages
...the signal number is not recognized
sigaction(2) , the string Unknown signal
is produced. The psiginfo function produces
the same output as the psignal function,
only it uses the signal number information
from the si argument. The message strings can...

@@
